---
title: ggplot2 2.2.0
output:
  html_document:
    self_contained: false
    highlight: NULL
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(forcats)

knitr::opts_chunk$set(
  fig.asp = 1 / 1.6,
  out.width = "75%",
  fig.width = 5,
  collapse = TRUE,
  comment = "#>",
  dpi = 96,
  fig.retina = NULL
)
```

I'm planning to release ggplot2 2.2.0 in early November. In preparation, I'd like to announce that a release candidate is now available: version 2.1.0.9001. Please try it out, and file an [issue on GitHub](https://github.com/tidyverse/ggplot2/issues) if you discover any problems. I hope we can find and fix any major issues before the official release.

Install the pre-release version with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("tidyverse/ggplot2")
```

If you discover a major bug that breaks your plots, please [file a minimal reprex](https://github.com/tidyverse/ggplot2/issues), and then roll back to the released version with:

```{r, eval = FALSE}
install.packages("ggplot2")
```

ggplot2 2.2.0 will be a relatively major release including:

* Subtitles and captions.
* A large rewrite of the facetting system.
* Improved theme options.
* Better stacking
* [Numerous bug fixes and minor improvements][news.md].

The majority of this work was carried out by [Thomas Pederson](https://github.com/thomasp85), who I was lucky to have as my "ggplot2 intern" this summer. Make sure to check out other visualisation packages: [ggraph](https://github.com/thomasp85/ggraph), [ggforce](https://github.com/thomasp85/ggforce), and [tweenr](https://github.com/thomasp85/tweenr).

## Subtitles and captions

Thanks to [Bob Rudis](https://rud.is), you can now add subtitles and captions:

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE, method = "loess") +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )
```

These are controlled by the theme settings `plot.subtitle` and `plot.caption`.

The plot title is now aligned to the left by default. To return to the previous centering, use `theme(plot.title = element_text(hjust = 0.5))`.

## Facets

The facet and layout implementation has been moved to ggproto and received a large rewrite and refactoring. This will allow others to create their own facetting systems, as descrbied in the *Extending ggplot2* vignette. Along with the rewrite a number of features and improvements has been added, most notably:
  
  * Functions in facetting formulas, thanks to 
    [Dan Ruderman](https://github.com/DanRuderman).
    
    ```{r}
    ggplot(diamonds, aes(carat, price)) + 
      geom_hex(bins = 20) + 
      facet_wrap(~cut_number(depth, 6))
    ```

  * Axes were dropped when the panels in `facet_wrap()` did not completely 
    fill the rectangle. Now, an axis is drawn underneath the hanging panels:
    
    ```{r}
    ggplot(mpg, aes(displ, hwy)) + 
      geom_point() + 
      facet_wrap(~class)
    ```

  * It is now possible to set the position of the axes through the `position` 
    argument in the scale constructor:
    
    ```{r}
    ggplot(mpg, aes(displ, hwy)) + 
      geom_point() + 
      scale_x_continuous(position = "top") + 
      scale_y_continuous(position = "right")
    ```

  * You can display a secondary axis that is a one-to-one transformation of the 
    primary axis with the `sec.axis` argument:
    
    ```{r}
    ggplot(mpg, aes(displ, hwy)) + 
      geom_point() + 
      scale_y_continuous(
        "mpg (US)", 
        sec.axis = sec_axis(~ . * 1.20, name = "mpg (UK)")
      )
    ```

  * Strips can be placed on any side, and the placement with respect to axes
    can be controlled with the `strip.placement` theme option.
    
    ```{r}
    ggplot(mpg, aes(displ, hwy)) + 
      geom_point() + 
      facet_wrap(~ drv, strip.position = "bottom") + 
      theme(
        strip.placement = "outside",
        strip.background = element_blank(),
        strip.text = element_text(face = "bold")
      ) +
      xlab(NULL)
    ```

## Theming

*   Blank elements can now be overridden again so you get the expected 
    behavior when setting e.g. `axis.line.x`. 
    
*   `element_line()` gets an `arrow` argument that lets you put arrows on axes.
    
    ```{r}
    arrow <- arrow(length = unit(0.4, "cm"), type = "closed")
    
    ggplot(mpg, aes(displ, hwy)) + 
      geom_point() + 
      theme_minimal() + 
      theme(
        axis.line = element_line(arrow = arrow)
      )
    ```

*   Control of legend styling has been improved. The whole legend area can be 
    aligned according to the plot area and a box can be drawn around all legends:
    
    ```{r}
    ggplot(mpg, aes(displ, hwy, shape = drv, colour = fl)) + 
      geom_point() + 
      theme(
        legend.justification = "top", 
        legend.box.margin = margin(3, 3, 3, 3, "mm"), 
        legend.box.background = element_rect(colour = "grey50")
      )
    ```

*   `panel.margin` and `legend.margin` have been renamed to `panel.spacing` 
    and `legend.spacing` respectively as this better indicates their roles.
    A new `legend.margin` has been actually controls the margin around each 
    legend.

*   When computing the height of titles ggplot2, now inclues the height of the
    descenders (i.e. the bits `g` and `y` that hang underneath). This makes 
    improves the margins around titles, particularly the y axis label. I have 
    also very slightly increased the inner margins of axis titles, and removed 
    the outer margins. 

*   The default themes has been tweaked by 
    [Jean-Olivier Irisson](http://www.obs-vlfr.fr/~irisson/) making them better
    match `theme_grey()`.

*   Lastly, the `theme()` function now has named arguments so autocomplete
    and documentation suggestions are vastly improved.

## Stacking bars
  
`position_stack()` and `position_fill()` now stack values in the reverse 
order of the grouping, which makes the default stack order match the legend.

```{r}
avg_price <- diamonds %>% 
  group_by(cut, color) %>% 
  summarise(price = mean(price)) %>% 
  ungroup() %>% 
  mutate(price_rel = price - mean(price))

ggplot(avg_price) + 
  geom_col(aes(x = cut, y = price, fill = color))
```

(Note also the new `geom_col()` which is short-hand for 
`geom_bar(stat = "identity")`, contributed by Bob Rudis.)

Additionally, you can now stack negative values:

```{r}
ggplot(avg_price) + 
  geom_col(aes(x = cut, y = price_rel, fill = color))
```

The overall ordering cannot necessarily be matched in the presence of 
negative values, but the ordering on either side of the x-axis will match. 

If you want to stack in the opposite order, try
[`forcats::fct_rev()`](http://forcats.tidyverse.org/reference/fct_rev.html):

```{r}
ggplot(avg_price) + 
  geom_col(aes(x = cut, y = price, fill = fct_rev(color)))
```

[news.md]: https://github.com/tidyverse/ggplot2/blob/master/NEWS.md
