---
title: "Extending ggplot2"
author: "Hadley Wickham"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggplot2)
```

This page aims at presenting the major steps involved in extending ggplot2 by:

* Creating a new geom, stat, or position adjustment
* Creating a new theme

This vignette documents the official extension mechanism provided in ggplot2 1.1.0. This vignette is a high-level adjunct to the low-level details found in `?Stat`, `?Geom` and `?theme`.

As you read this document, you'll see many things that will scratch your head and make you wonder, why on earth is it designed that way? Mostly it's historical accident - I wasn't a terribly good R programmer when I started writing ggplot2 and I made a lot of questionable decisions. We cleaned up as many as possible in the 1.1.0 release, but some fixes simply weren't worth the effort.

## ggproto

All ggplot2 objects are built using the ggproto system of object oriented programming. This OO system is used only in one place: ggplot2. This is mostly historical accident: ggplot2 started off using [proto](https://cran.r-project.org/web/packages/proto) because I needed mutable objects. This was well before the creation of (the briefly lived) [mutatr](http://vita.had.co.nz/papers/mutatr.html), reference classes and R6: proto was the only game in town.

But why ggproto? Well when we turned to add an official extension mechanism to ggplot2, we found a major problem that caused problems when proto objects were extended in a different package (methods were evaluated in ggplot2, not the package where the extension was added). We tried converting to R6, but it was a poor fit for the needs of ggplot2. We could've modified proto, but that would've first involved understand exactly how proto worked, and secondly making sure that the changes didn't affect other users of proto.

It's strange to say, but this is a case where inventing a new OO system was actually the right answer to the problem! Fortunately Winston is now very good at creating OO systems, so it only took him a day to come up with ggproto: it maintains all the features of proto that ggplot2 needs, while allowing cross package inheritance to work.

Here's a quick demo of ggproto in action:

```{r}
A <- ggproto("A", NULL,
  x = 1,
  inc = function(self) {
    self$x <- self$x + 1
  }
)
A$x
A$inc()
A$x
A$inc()
A$inc()
A$x
```

The majority of ggplot2 classes are immutable and static: the methods neither use nor modify state in the class. They're mostly used as a convenient way of bundling related methods together.

To create a new geom or stat, you will just create a new ggproto that inherits from `Stat`, `Geom` and override the methods described below.

## Creating a new stat

Start with the simplest stat. It just picks some of the rows:

```{r}
StatChull <- ggproto("StatChull", Stat,
  compute_group = function(data, scales) {
    
    data[chull(data$x, data$y), , drop = FALSE]
  },
  
  required_aes = c("x", "y")
)

stat_chull <- function(mapping = NULL, data = NULL, geom = "polygon",
                       position = "identity", show.legend = NA, 
                       inherit.aes = TRUE, ...) {
  layer(
    stat = StatChull, data = data, mapping = mapping, geom = geom, 
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(...)
  )
}

ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  stat_chull(fill = NA, colour = "black")
```

All wrapper functions follow the same form - you specify defaults in the function arguments and then call the `layer()` function, sending `...` into the `params` argument. The arguments in `...` will either be arguments for the geom (if you're making a stat function), or aesthetics that we're setting. `layer()` takes care of teasing them apart and making sure they're stored in the right place.

Note that ggplot takes care of adding constant aesthetic variables within the group. So this works without you having to do anything:

```{r}
ggplot(mpg, aes(displ, hwy, colour = drv)) + 
  geom_point() + 
  stat_chull(fill = NA)
```

In an ideal world, I would have originally called these things layers, so you'd use `layer_chull()` instead of `

You might also consider createing a `geom_chull()`:

```{r}
geom_chull <- function(mapping = NULL, data = NULL, 
                       position = "identity", show.legend = NA, 
                       inherit.aes = TRUE, ...) {
  layer(
    stat = StatChull, data = data, mapping = mapping,
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(...)
  )
}
```

This doesn't allow you to use different geoms with the stat, but that seems appropriate here since the convex hull is primarily a polygonal feature.

A more realistic stat will actually do some computation. Let's implement a simple version of `geom_smooth()` that adds a line of best fit to a plot.

```{r}
StatLm <- ggproto("StatLm", Stat, 
  required_aes = c("x", "y"),
  
  compute_group = function(data, scales, params) {
    rng <- range(data$x, na.rm = TRUE)
    grid <- data.frame(x = rng)
    
    mod <- lm(y ~ x, data = data)
    grid$y <- predict(mod, newdata = grid)
    
    grid
  }
)

stat_lm <- function(mapping = NULL, data = NULL, geom = "line",
                    position = "identity", show.legend = NA, 
                    inherit.aes = TRUE, ...) {
  layer(
    stat = StatLm, data = data, mapping = mapping, geom = geom, 
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(...)
  )
}

ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  stat_lm()
```

Let's extend this to allow the user to supply a formula and the number of points used to generate the grid.

```{r}
StatLm <- ggproto("StatLm", Stat, 
  required_aes = c("x", "y"),
  
  compute_group = function(data, scales, params, n = 100, formula = y ~ x) {
    rng <- range(data$x, na.rm = TRUE)
    grid <- data.frame(x = seq(rng[1], rng[2], length = n))
    
    mod <- lm(formula, data = data)
    grid$y <- predict(mod, newdata = grid)
    
    grid
  }
)

stat_lm <- function(mapping = NULL, data = NULL, geom = "line",
                    position = "identity", show.legend = NA, 
                    inherit.aes = TRUE, n = 50, formula = y ~ x, 
                    ...) {
  layer(
    stat = StatLm, data = data, mapping = mapping, geom = geom, 
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(n = n, formula = formula, ...)
  )
}

ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  stat_lm(formula = y ~ poly(x, 3), n = 5)
```

## Creating a new geom

It's easiest to start with a simple example. The code below creates a simplified version of `geom_point()`:

```{r}
GeomSimplePoint <- ggproto("GeomSimplePoint", Geom,
  required_aes = c("x", "y"),
  default_aes = aes(shape = 19, colour = "black"),
  draw_key = draw_key_point,

  draw_panel = function(data, panel_scales, coord) {
    coords <- coord$transform(data, panel_scales)
    grid::pointsGrob(
      coords$x, coords$y,
      pch = coords$shape,
      gp = grid::gpar(col = coords$colour)
    )
  }
)

geom_simple_point <- function(mapping = NULL, data = NULL, stat = "identity",
                              position = "identity", show.legend = NA, 
                              inherit.aes = TRUE, ...) {
  layer(
    geom = GeomSimplePoint, mapping = mapping,  data = data, stat = stat, 
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(...)
  )
}

ggplot(mpg, aes(displ, hwy)) + geom_simple_point()
```

Much should look familiar from creating new scales. There are two key pieces:

* A `GeomPoint2` object that defines the behaviour of the plot.
* A `geom_point2()` wrapper function that creates a layer.

The object is more complicated. You always need to provide fields/methods for the four pieces shown above:

* `required_aes` is a character vector which lists all the aesthetics that
  the user must provide.
  
* `default_aes` lists the aesthetics that have default values.

* `draw_key` provides the function used to draw the key in the legend. 
  You can see a list of all the build in key functions in `?draw_key`
  
* `draw_group()` is where the magic happens. This function takes three
  arguments and returns a grid grob. It is called once for each panel.
  It's the most complicated part and is described in more detail below.
  
`draw_group()` has three arguments:

* `data`: a data frame with one column for each aesthetic.

* `panel_scales`: a list containing information about the x and y scales
  for the current panel.

* `coord`: an object describing the coordinate system.

Generally you won't use `panel_scales` and `coord` directly, but you will always use them to transform the data: `coords <- coord$transform(data, panel_scales)`. This creates a data frame where position variables are scaled to the range 0--1. You then take this data and call a grid grob function.

ggplot2 is built on top of grid, so you'll need to know the basics of drawing with grid. If you're serious about adding a new geom, I'd recommend buying [R graphics](http://amzn.com/B00I60M26G) by Paul Murrell. It tells you everything you need to know about drawing with grid.

Overriding `draw_panel()` is most appropriate if there is one graphic element per row. In other cases, you'll get one row per group. For example, take polygons. The following code is a somewhat simplified version of `GeomPolygon`:

```{r}
GeomSimplePolygon <- ggproto("GeomPolygon", Geom,
  required_aes = c("x", "y"),
  
  default_aes = aes(colour = "NA", fill = "grey20", size = 0.5, linetype = 1, alpha = 1),

  draw_key = draw_key_polygon,

  draw_group = function(data, panel_scales, coord) {
    n <- nrow(data)
    if (n <= 2) return(grid::nullGrob())

    coords <- coord$transform(data, panel_scales)
    # A polygon can only have a single colour, fill, etc, so take from first row
    first_row <- coords[1, , drop = FALSE]

    grid::polygonGrob(
      coords$x, coords$y, 
      default.units = "native",
      gp = grid::gpar(
        col = first_row$colour,
        fill = scales::alpha(first_row$fill, first_row$alpha),
        lwd = first_row$size * .pt,
        lty = first_row$linetype
      )
    )
  }
)
geom_simple_polygon <- function(mapping = NULL, data = NULL, stat = "chull",
                                position = "identity", show.legend = NA, 
                                inherit.aes = TRUE, ...) {
  layer(
    geom = GeomSimplePolygon, mapping = mapping, data = data, stat = stat, 
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(...)
  )
}

ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  geom_simple_polygon(aes(colour = class), fill = NA)
```

There are a few things to note here:

* We override `draw_group()` instead of `draw_layer()` because we want
  one polygon per group, not one polygon per row. If you look at the source
  code for the original `GeomPolygon` you'll see it actually overrides
  `geom_layer()` because it uses some tricks to make `polygonGrob()` produce
  multiple polygons in one call. This is considerably more complicated, but 
  gives better performance.
  
* If the data contains two or fewer points, there's no point trying to draw
  a polygon, so we return a `nullGrob()`. This is the graphical equivalent
  of `NULL`: it's a grob that doesn't draw anything and doesn't take up
  any space.
  
* Note the units: `x` and `y` should always be drawn in "native" units. 
  (The default units for `pointGrob()` is a native, so we didn't need to 
  change it there). `lwd` is measured in points, but ggplot2 uses mm, 
  so we need to multiply it by the adjustment factor `.pt`.

## Creating your own theme

Instead of modifying an existing theme piecemeal, you might want to start from scratch and create your own unique theme.
If you want to create your complete theme, you'll need to understand a bit more about the structure of themes.


### Global settings

There are four elements that affect the global appearance of the plot.

Element      | Theme function    | Description
-------------|-------------------|------------------------
line         | `element_line()`  | all line elements
rect         | `element_rect()`  | all rectangular elements
text         | `element_text()`  | all text
title        | `element_text()`  | all text in title elements (plot, axes & legend)

These set default properties that are inherited by more specific settings. These are most useful for setting an overall "background" colour and overall font settings (e.g. family and size).

```{r axis-line-ex}
df <- data.frame(x = 1:3, y = 1:3)
base <- ggplot(df, aes(x, y)) + geom_point()

base + theme(text = element_text(colour = "pink"))
```

It's also possible to use `rel()` to specify a relative size.

Inheritance

Complete vs. incomplete.

`%+replace%`

### Complete vs incomplete

It is useful to understand the difference between complete and incomplete theme objects. A *complete* theme object is one produced by calling a theme function with the attribute `complete = TRUE`. 

Theme functions `theme_grey()` and `theme_bw()` are examples of complete theme functions. Calls to `theme()` produce *incomplete* theme objects, since they represent (local) modifications to a theme object rather than returning a complete theme object per se. When adding an incomplete theme to a complete one, the result is a complete theme. 

Complete and incomplete themes behave somewhat differently when added to a ggplot object:

* Adding an incomplete theme augments the current theme object, replacing only 
  those properties of elements defined in the call to `theme()`. In particular, 
  any NULL element properties specified in `theme()`.
  
* Adding a complete theme wipes away the existing theme and applies. 

To deal with this problem, the concept of a complete theme (object) was introduced in the new theming system. In particular, `theme_grey` and `theme_bw` are both complete theme functions and return complete theme objects when called. When added to a plot, a complete theme object overrides the current default theme and in fact *replaces* it. This idea is implemented by endowing a theme function with the attribute `"complete"`: its value is `TRUE` for complete themes and `FALSE` for incomplete themes. Examples are shown in the next section.

### Modifying a theme function

There is a second type of inheritance in themes

```{r}
theme_grey()$legend.key

new_theme <- theme_grey() + theme(legend.key = element_rect(colour = "red"))
new_theme$legend.key

new_theme <- theme_grey() %+replace% theme(legend.key = element_rect(colour = "red"))
new_theme$legend.key
```
