---
title: "Profiling Performance"
author: "Thomas Lin Pedersen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profiling Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In order to continuously monitor the performance of ggplot2
the following piece of code is used to generate a profile and inspect it:

```{r}
library(ggplot2)
library(profvis)

p <- ggplot(mtcars, aes(x = mpg, y = disp)) + 
  geom_point() + 
  facet_grid(gear ~ cyl)

profile <- profvis(for (i in seq_len(100)) ggplotGrob(p))

profile
```

```{r, eval=FALSE, include=FALSE}
saveRDS(profile, file.path('profilings', paste0(packageVersion('ggplot2'), '.rds')))
```

In general, a minimal plot is used so that profiles are focused on low-level,
general code, rather than implementations of specific geoms. This might be
expanded at the point where improving performance of specific geoms becomes a
focus. Further, the profile focuses on the steps up until a final gtable have
been constructed. Any performance problems in rendering is likely due to grid 
and the device, more than ggplot2.

Profiles for old version are kept for reference and can be accessed at 
the [github repository](https://github.com/tidyverse/ggplot2/tree/master/vignettes/profilings).
Care should be taken in not comparing profiles across versions, as 
changes to code outside of ggplot2 can have profound effect on the results.
Thus, the intend of profiling is to identify bottlenecks in the implementation
that are ripe for improvement, more then to quantify improvements to performance
over time.

## Performance focused changes across versions
To keep track of changes focused on improving the performance of gtable they
are summarised below:

### v`r packageVersion('ggplot2')`
