---
title: "Profiling Performance"
author: "Thomas Lin Pedersen"
output: rmarkdown::html_vignette
description: |
  Monitoring the performance of your plots.
vignette: >
  %\VignetteIndexEntry{Profiling Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In order to continuously monitor the performance of ggplot2
the following piece of code is used to generate a profile and inspect it:

```{r}
library(ggplot2)
library(profvis)

p <- ggplot(mtcars, aes(x = mpg, y = disp)) + 
  geom_point() + 
  facet_grid(gear ~ cyl)

profile <- profvis(for (i in seq_len(100)) ggplotGrob(p))

profile
```

```{r, eval=FALSE, include=FALSE}
saveRDS(profile, file.path('profilings', paste0(packageVersion('ggplot2'), '.rds')))
```

In general, a minimal plot is used so that profiles are focused on low-level,
general code, rather than implementations of specific geoms. This might be
expanded at the point where improving performance of specific geoms becomes a
focus. Further, the profile focuses on the steps up until a final gtable have
been constructed. Any performance problems in rendering is likely due to grid 
and the device, more than ggplot2.

Profiles for old version are kept for reference and can be accessed at 
the [github repository](https://github.com/tidyverse/ggplot2/tree/master/vignettes/profilings).
Care should be taken in not comparing profiles across versions, as 
changes to code outside of ggplot2 can have profound effect on the results.
Thus, the intend of profiling is to identify bottlenecks in the implementation
that are ripe for improvement, more then to quantify improvements to performance
over time.

## Performance focused changes across versions
To keep track of changes focused on improving the performance of gtable they
are summarised below:

### v`r packageVersion('ggplot2')`
- **Avoid costly factor construction in scale_apply** This issue only really 
  appeared when plotting huge datasets (>1e6 rows). In order to train and map 
  the position aesthetics rows has to be matched based on their panel. This 
  require splitting the row indexes by panel which included a `factor()` call. 
  The factor constructor is very slow for large vectors and can be simplified 
  considerably for this specific case. Further, the split can be avoided 
  completely when there is only one panel

### v3.1.0
- **Caching of calls to `grid::descentDetails()`** The absolute biggest offender
  was the construction of titles. In recent versions this has included calls to
  `grid::descentDetails()` to ensure that they are aligned across plots, but 
  this is quite heavy. These calls are now cached so they only have to be 
  calculated once per font setting.
- **Use a performant `data.frame` constructor throughout the codebase** The
  `data.frame()` function carries a lot of overhead in order to sanitize and 
  check the input. This is generally not needed if you are sure about the input
  and will just lead to slower code. The `data.frame()` call is now only used
  when dealing with output from other packages where the extra safety is a 
  benefit.
- **Use a performant alternative to `utils::modifyList`** `modifyList()` is a 
  nice convenience function but carries a lot of overhead. It was mainly used
  in the plot element constructions where it slowed down the application of 
  theme settings. A more performant version has been added and used throughout.
- **Speed up position transformation** The `transform_position` helper was 
  unreasonably slow due to the slowness of getting and assigning columns in 
  data.frame. The input is now treated as a list during transformation.
