---
title: "FAQ: Customising"
---

```{=html}
<style>
  .content h3 {
    margin-top: -30px !important;
  }
  
  details {
    margin-bottom: 40px;
  }
</style>
```
```{r, include = FALSE}
library(ggplot2)
library(tibble)
knitr::opts_chunk$set(
  fig.dpi = 300, 
  collapse = TRUE, 
  comment = "#>",
  fig.asp = 0.618,
  fig.width = 6,
  out.width = "80%"
  )
```

## Legends

### How can I change the legend title?

Change the label for the aesthetic the legend is drawn for in `labs()`.

<details>

<summary>See example</summary>

By default your legend label will be the name of the variable that is mapped to the aesthetic the legend is drawn for.
You can change the title of your legend using `labs()`.

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The points are coloured by three
#|  types of drive train, which is displayed in a legend with the title 'Drive
#|  train'."
ggplot(mpg, aes(x = hwy, y = cty, color = drv)) +
  geom_point() +
  labs(color = "Drive train")
```

If a legend is drawn for multiple aesthetics, you'll want to update the title for all of them.

```{r}
#| fig.alt:
#| - "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The point shapes and colours 
#|  indicate three types of drive train. The shapes and colours are displayed in
#|  separate legends titled 'drv' and 'Drive train' respectively."
#| - "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The point shapes and colours 
#|  indicate three types of drive train. The shapes and colours are displayed in
#|  a single legend titled 'Drive train'."
#|
# not this
ggplot(mpg, aes(x = hwy, y = cty, color = drv, shape = drv)) +
  geom_point() +
  labs(color = "Drive train")

# but this
ggplot(mpg, aes(x = hwy, y = cty, color = drv, shape = drv)) +
  geom_point() +
  labs(color = "Drive train", shape = "Drive train")
```

</details>

### How can I increase the spacing between legend keys?

Increase the horizontal space between legend keys with `legend.spacing.x` in `theme()`.
This argument takes a unit object created with `grid::unit()`.

<details>

<summary>See example</summary>

If you have a horizontal legend, generally placed on top or bottom of the plot with `legend.position = "top"` or `"bottom"`, you can change the spacing between legend keys with `legend.spacing.x`.
You can supply a unit object to this argument, e.g. `unit(1.0, "cm")` for 1 cm space between legend keys.
See the documentation for `grid::unit()` for more options for units.

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The points are coloured by three
#|  types of drive train, which is displayed in a legend at the bottom of the 
#|  plot in a horizontal orientation. Legend elements are spaced widely apart."
ggplot(mpg, aes(x = hwy, y = cty, color = drv)) +
  geom_point() +
  theme(
    legend.position = "bottom", 
    legend.spacing.x = unit(1.0, "cm")
    )
```

For vertical legends changing `legend.spacing.y` changes the space between the legend title and the keys, but not between the keys, e.g. see the large space between the legend title and keys.

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The points are coloured by three
#|  types of drive train, which is displayed in a legend at the right of the 
#|  plot. In the legend, there is a large space between the title and keys."
ggplot(mpg, aes(x = hwy, y = cty, color = drv)) +
  geom_point() +
  theme(legend.spacing.y = unit(3.0, "cm"))
```

In order to change the space between the legend keys, you can first make the key size bigger with `legend.key.size` and then remove the grey background color with `legend.key`.

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The points are coloured by three
#|  types of drive train, which is displayed in a legend at the right of the 
#|  plot. In the legend, elements are placed widely apart and the title is
#|  to the left of the keys."
ggplot(mpg, aes(x = hwy, y = cty, color = drv)) +
  geom_point() +
  theme(
    legend.key.size = unit(1.5, "cm"),
    legend.key = element_rect(color = NA, fill = NA)
    )
```

Note that the legend title is no longer aligned with the keys with this approach.
You can also shift it over with the `hjust` setting of `legend.title`.

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The points are coloured by three
#|  types of drive train, which is displayed in a legend at the right of the 
#|  plot. In the legend, elements are placed widely apart and the title is
#|  aligned with the keys."
ggplot(mpg, aes(x = hwy, y = cty, color = drv)) +
  geom_point() +
  theme(
    legend.key.size = unit(1.5, "cm"),
    legend.key = element_rect(color = NA, fill = NA),
    legend.title = element_text(hjust = 0.5)
  )
```

</details>

### How can I change the key labels in the legend?

If you don't want to change the levels of the variable the legend is being drawn for, you can change the key labels at the time of drawing the plot using the `labels` argument in the appropriate `scale_*()` function, e.g. `scale_colour_discrete()` if the legend is for a discrete variable mapped to the fill aesthetic.

<details>

<summary>See example</summary>

The `labels` argument of `scale_*` functions takes named vectors, which what we would recommend using for relabeling keys in a legend.
Using named lists allows you to declare explicitly which label is assigned to which level, without having to keep track of level order.

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The points are coloured by three
#|  types of drive train, which is displayed in a legend at the right of the 
#|  plot. The legend items are name '4-wheel drive', 'Front-wheel drive' and
#|  'Rear-wheel drive' from top to bottom."
ggplot(mpg, aes(x = hwy, y = cty, color = drv)) +
  geom_point() +
  scale_color_discrete(
    labels = c("4" = "4-wheel drive", 
               "f" = "Front-wheel drive", 
               "r" = "Rear-wheel drive")
    )
```

</details>

### How can I change the font sizes in the legend?

Set your preference in `legend.text` for key labels and `legend.title` in `theme()`.
In both cases, set font size in the `size` argument of `element_text()`, e.g. `legend.text = element_text(size = 14)`.

<details>

<summary>See example</summary>

Font characteristics of a legend can be controlled with the `legend.text` and `legend.title` elements of `theme()`.
You can use the following for 14 pts text for legend key labels and 10 pts text for legend title.
(Note that this doesn't result in a visually pleasing legend, by default ggplot2 uses a larger font size for the legend title than the legend text.)

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The points are coloured by seven 
#|  types of cars, which is displayed in the legend on the right of the plot.
#|  The labels in the legend have a larger font size than the title."
ggplot(mpg, aes(x = hwy, y = cty, color = class)) +
  geom_point() +
  theme(
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 10)
  )
```

For further customization of legend text, see the documentation for `element_text()`, e.g. you can change font colors or font face as well.

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The points are coloured by seven 
#|  types of cars, which is displayed in the legend on the right of the plot.
#|  The labels in the legends have a large, red font. The title has a smaller,
#|  black font and is in bold and italic."
ggplot(mpg, aes(x = hwy, y = cty, color = class)) +
  geom_point() +
  theme(
    legend.text = element_text(size = 14, color = "red"),
    legend.title = element_text(size = 10, face = "bold.italic")
  )
```

</details>

## Colours

### How can I change the background colour of plot?

Set the color in `panel.background` element of `theme()` with `element_rect()`, which takes arguments like `fill` (for background fill color) and `colour` (for background border color.

<details>

<summary>See example</summary>

You can set the background colour of the plot with `panel.background` in `theme()`.
In the following example the border is made thicker with `linewidth = 3` to

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The panel background of the plot
#|  is light blue and is outlined in red with a thick stroke."
ggplot(mpg, aes(x = hwy, y = cty)) +
  geom_point() +
  theme(panel.background = element_rect(fill = "lightblue", colour = "red", linewidth = 3))
```

If you want to change the colour of the plotting area but not the panel where the panel, you can so the same thing with `plot.background`.

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The plot background is light blue 
#|  and is outlined in red with a thick stroke. The panel background remains 
#|  grey."
ggplot(mpg, aes(x = hwy, y = cty)) +
  geom_point() +
  theme(plot.background = element_rect(fill = "lightblue", colour = "red", linewidth = 3))
```

Note that ggplot2 has a variety of [complete themes](https://ggplot2.tidyverse.org/reference/ggtheme.html) that might already do what you're hoping to accomplish.
For example, if you prefer a more minimal look to your plots, without the grey background, you might try `theme_minimal()`.

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. There is no visible panel 
#|  background and grid lines are in light grey."
ggplot(mpg, aes(x = hwy, y = cty)) +
  geom_point() +
  theme_minimal()
```

And you can continue customization based on one of these themes.

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. There is no visible panel 
#|  background and grid lines are in light grey. The plot as a whole is outlined
#|  by a thick red line."
ggplot(mpg, aes(x = hwy, y = cty)) +
  geom_point() +
  theme_minimal() +
  theme(plot.background = element_rect(colour = "red", linewidth = 3))
```

You might also find the [**thematic**](https://rstudio.github.io/thematic/) package useful for simplified theming of your plots.

</details>

### How can I change the colour NAs are represented with in a plot?

You can set the color of `NA` with the `na.value` argument in the appropriate `scale_*()` function, e.g. `scale_fill_discrete(na.value = "purple")` to make `NA`s purple.

<details>

<summary>See example</summary>

Suppose you have the following data frame with two discrete variables, one of which has an `NA`.

```{r}
df <- tibble::tribble(
  ~group, ~outcome,
  1,      "yes",
  1,      "no",
  2,      "yes",
  2,      "no",
  2,      "no",
  2,      NA 
)
```

By default, ggplot2 uses grey to represent `NA`s.

```{r}
#| fig.alt: "A stacked bar chart showing two groups on the x-axis and counts
#|  on the y-axis. Within a stacked bar, two different outcomes and 'NA's are 
#|  distinguished by fill colour."
ggplot(df, aes(x = group, fill = outcome)) +
  geom_bar()
```

You can change the color of `NA` with `scale_fill_discrete()` in this case, e.g. make it purple.

```{r}
#| fig.alt: "A stacked bar chart showing two groups on the x-axis and counts
#|  on the y-axis. Within a stacked bar, two different outcomes and 'NA's are 
#|  distinguished by fill colour. The 'NA' fill colour is purple."
ggplot(df, aes(x = group, fill = outcome)) +
  geom_bar() +
  scale_fill_discrete(na.value = "purple")
```

You can also set the color to `"transparent"`.
In the plot below this is shown with `theme_minimal()` to demonstrate how that looks on a plot with a transparent background.
Note that while this is possible, setting the colour to transparent as such wouldn't be recommended in this particular case as it gives the appearance of a floating bar.

```{r}
#| fig.alt: "A stacked bar chart showing two groups on the x-axis and counts
#|  on the y-axis. Within a stacked bar, two different outcomes and 'NA's are 
#|  distinguished by fill colour. The 'NA' fill colour is transparent, giving
#|  the appearance that one of the stacked bars is floating."
ggplot(df, aes(x = group, fill = outcome)) +
  geom_bar() +
  scale_fill_discrete(na.value = "transparent") +
  theme_minimal()
```

</details>

## Fonts

### How can I change the default font size in ggplot2?

Set `base_size` in the theme you're using, which is `theme_gray()` by default.

<details>

<summary>See example</summary>

The base font size is 11 pts by default.
You can change it with the `base_size` argument in the theme you're using.
See the [complete theme documentation](https://ggplot2.tidyverse.org/reference/ggtheme.html) for more high level options you can set.

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The points are coloured by seven
#|  types of car. All text sizes in the axes and legend are large."
ggplot(mpg, aes(x = hwy, y = cty, color = class)) +
  geom_point() +
  theme_gray(base_size = 18)
```

If you would like all plots within a session/document to use a particular base size, you can set it with `set_theme()`.
Run the following at the beginning of your session or include on top of your R Markdown document.

```{r eval = FALSE}
set_theme(theme_gray(base_size = 18))
```

</details>

### How can I change the font size of the plot title and subtitle?

Set your preference in `plot.title` and `plot.subtitle` in `theme()`.
In both cases, set font size in the `size` argument of `element_text()`, e.g. `plot.title = element_text(size = 20)`.

<details>

<summary>See example</summary>

Font characteristics of plot titles and subtitles can be controlled with the `plot.title` and `plot.subtitle` elements of `theme()`.
You can use the following for 20 pts text for the plot title and 15 pts text for the plot subtitle.

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The plot has a large title 
#|  displaying 'This is the plot title' and a less large subtitle displaying 
#|  'And this is the subtitle' at the top of the plot."
ggplot(mpg, aes(x = hwy, y = cty)) +
  geom_point() +
  labs(
    title = "This is the plot title",
    subtitle = "And this is the subtitle"
  ) +
  theme(
    plot.title = element_text(size = 20),
    plot.subtitle = element_text(size = 15)
  )
```

For further customization of plot title and subtitle, see the documentation for `element_text()`, e.g. you can change font colors or font face as well.

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The plot has a large red title 
#|  displaying 'This is the plot title' and a less large subtitle in bold and 
#|  italic displaying 'And this is the subtitle' at the top of the plot."
ggplot(mpg, aes(x = hwy, y = cty)) +
  geom_point() +
  labs(
    title = "This is the plot title",
    subtitle = "And this is the subtitle"
  ) +
  theme(
    plot.title = element_text(size = 20, color = "red"),
    plot.subtitle = element_text(size = 15, face = "bold.italic")
  )
```

</details>

### How can I change the font size of axis labels?

Set your preference in `axis.title`.
`axis.title.x`, or `axis.title.y` in `theme()`.
In both cases, set font size in the `size` argument of `element_text()`, e.g. `axis.text = element_text(size = 14)`.

<details>

<summary>See example</summary>

Font characteristics of axis labels can be controlled with `axis.title.x` or `axis.title.y` (or `axis.title` if you the same settings for both axes).

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The x-axis title displays 
#|  'This is HUGE' in a large font size, and the y-axis title displays 
#|  'This is small' in a smaller font size."
ggplot(mpg, aes(x = hwy, y = cty)) +
  geom_point() +
  labs(
    x = "This is HUGE",
    y = "This is small"
  ) +
  theme(
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 10)
  )
```

For further customization of plot title and subtitle, see the documentation for `element_text()`, e.g. you can change font colors or font face as well.

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. The x-axis title displays 
#|  'This is HUGE' in a large, red font, and the y-axis title displays 
#|  'This is tiny' in a smaller, bold and italic font."
ggplot(mpg, aes(x = hwy, y = cty)) +
  geom_point() +
  labs(
    x = "This is HUGE",
    y = "This is tiny"
  ) +
  theme(
    axis.title.x = element_text(size = 20, color = "red"),
    axis.title.y = element_text(size = 10, face = "bold.italic")
  )
```

You can also change the size of the axis text (e.g. numbers at the axis ticks) using `axis.text` (or `axis.text.x` and `axis.text.y` if you want to set different sizes).

```{r}
#| fig.alt: "A scatter plot showing the highway miles per gallon on the x-axis
#|  and city miles per gallon on the y-axis. Both the x and the y axis titles
#|  display 'The axis labels are the same size' in a large font. Both axis 
#|  labels are displayed in a larger, blue font."
ggplot(mpg, aes(x = hwy, y = cty)) +
  geom_point() +
  labs(
    x = "The axis labels are the same size",
    y = "The axis labels are the same size"
  ) +
  theme(
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 20, color = "blue")
  )
```

</details>

### What is the default size of `geom_text()` and how can I change the font size of `geom_text()`?

The default font size of `geom_text()` is about 3.87.

```{r}
get_geom_defaults(geom_text)$size
```

You can change the size using the `size` argument in `geom_text()` for a single plot. If you want to use the same updated size, you can set this with `update_geom_defaults()`, e.g. `update_geom_defaults("text", list(size = 6))`.

One tricky thing is that this `size` is in **mm** while the `size` of `element_text()` is in pt.
If you want to match `geom_text()`'s font size to the theme's size,
specify `size.unit = "pt"` in `geom_text()` for a single plot,
or divide the number by the `.pt` constant variable for `update_geom_defaults()`.
Please refer to ["Font size" section of the aesthetic specifications](https://ggplot2.tidyverse.org/articles/ggplot2-specs.html#font-size) for more details.

<details>

<summary>See example</summary>

Suppose you have the following data frame and visualization.

```{r}
#| fig.alt: "A plot showing text at diagonal positions with the labels 'two', 
#| 'three' and 'four'."
df <- tibble::tribble(
  ~x, ~y, ~name,
  2,  2,  "two",
  3,  3,  "three",
  4,  4,  "four"
)

ggplot(df, aes(x = x, y = y, label = name)) +
  geom_text()
```

You can set the size of the text with the following.

```{r}
#| fig.alt: "A plot showing larger text at diagonal positions with the labels 
#| 'two', 'three' and 'four'."
ggplot(df, aes(x = x, y = y, label = name)) +
  geom_text(size = 6)
```

Or you can map it to the `size` `aes`thetic. In the following size is determined by the `x` value with `scale_size_identity()`.

```{r}
#| fig.alt: "A plot showing text at diagonal positions with the labels 'two',
#|  'three' and 'four' that increase in size from left to right."
ggplot(df, aes(x = x, y = y, label = name)) +
  geom_text(aes(size = x)) +
  scale_size_identity()
```

If you want to use the same updated size for `geom_text()` in a series of plots in a session/R Markdown document, you can set use `update_geom_defaults()` to update the default size, e.g. if you want the size for all `geom_text()` to be 6, use `update_geom_defaults("text", list(size = 6))`.

</details>
