---
title: "FAQ: Barplots"
---

```{=html}
<style>
  .content h3 {
    margin-top: -30px !important;
  }
  
  details {
    margin-bottom: 40px;
  }
</style>
```
```{r, include = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

knitr::opts_chunk$set(
  fig.dpi = 300, 
  collapse = TRUE, 
  comment = "#>",
  fig.asp = 0.618,
  fig.width = 6,
  out.width = "80%"
  )
```

## Colors

### How can I change the color of the bars in my bar plot?

If using the same color for all bars, define the `fill` argument in `geom_bar()` (or `geom_col()`).
If assigning color based on another variable, map the variable to the `fill` `aes`thetic, and if needed, use one of the `scale_fill_*()` functions to set colors.

<details>

<summary>See example</summary>

You can set all bars to be a given color with the `fill` argument of `geom_bar()`.

```{r}
ggplot(mpg, aes(x = drv)) +
  geom_bar(fill = "blue")
```

Alternatively, if the colors should be based on a variable, this should be should happen in the `aes()` mapping.

```{r}
ggplot(mpg, aes(x = drv, fill = drv)) +
  geom_bar()
```

And if you want to then customize the colors, one option is `scale_fill_manual()`, which allows you to manually assign colors to each bar.
See other `scale_fill_*()` functions for more options for color choices.

```{r}
ggplot(mpg, aes(x = drv, fill = drv)) +
  geom_bar() +
  scale_fill_manual(values = c("purple", "orange", "darkblue"))
```

</details>

## Spacing and widths

### How can I increase the space between the bars in my bar plot?

Set the `width` of `geom_bar()` to a small value to obtain narrower bars with more space between them.

<details>

<summary>See example</summary>

By default, the `width` of bars is `0.9` (90% of the resolution of the data).
You can set this argument to a lower value to get bars that are narrower with more space between them.

```{r}
ggplot(mpg, aes(x = drv)) +
  geom_bar(width = 0.5)

ggplot(mpg, aes(x = drv)) +
  geom_bar(width = 0.1)
```

</details>

### How can I remove the space between the bars and the x-axis?

Adjust the `expand` argument in `scale_y_continuous()`, e.g. add `scale_y_continuous(expand = expansion(mult = c(0, 0.05)))` to remove the expansion on the lower end of the y-axis but keep the expansion on the upper end of the y-axis at 0.05 (the default expansion for continuous scales).

<details>

<summary>See example</summary>

By default ggplot2 expands the axes so the geoms aren't flush against the edges of the plot.

```{r}
ggplot(mpg, aes(x = drv)) +
  geom_bar()
```

To remove the spacing between the bars and the x-axis, but keep the spacing between the bars and the top of the plot, use the following.

```{r}
ggplot(mpg, aes(x = drv)) +
  geom_bar() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)))
```

To achieve the opposite, switch the values in `mult`.
Note that the tallest bar is now flush against top of the plot.

```{r}
ggplot(mpg, aes(x = drv)) +
  geom_bar() +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0)))
```

To adjust spacing around the x-axis, adjust the `expand` argument in `scale_x_discrete()`.
Note that this places the bars flush against the left side and leaves some space on the right side.

```{r}
ggplot(mpg, aes(x = drv)) +
  geom_bar() +
  scale_x_discrete(expand = expansion(add = c(0, 0.6)))
```

The default look of a bar plot can be achieved with the following.

```{r}
ggplot(mpg, aes(x = drv)) +
  geom_bar() +
  scale_x_discrete(expand = expansion(add = 0.6)) +
  scale_y_continuous(expand = expansion(mult = 0.05))
```

</details>

### How do I ensure that bars on a dodged bar plot have the same width?

Set `position = position_dodge2(preserve = "single")` in `geom_bar()`.

<details>

<summary>See example</summary>

In the following plot the bars have differing widths within each level of `drv` as there are differing levels of `class` represented.

```{r}
ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar(position = "dodge")
```

You can use `position_dodge2()` with `preserve = "single"` to address this.

```{r}
ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar(position = position_dodge2(preserve = "single"))
```

</details>

## Stacked bar plots

### How can I create a stacked bar plot displaying a conditional distribution where each stack is scaled to sum to 100%?

Use `position = "fill"` in `geom_bar()` or `geom_col()`.
If you also want to show percentages on the axis, use `scales::label_percent()`.

<details>

<summary>See example</summary>

The following plot is useful for comparing counts but not as useful for comparing proportions, which is what you need if you want to be able to make statements like "in this sample, it's more likely to have a two-seater car that has rear-wheel drive than an SUV that has rear-wheel drive".

```{r}
ggplot(mpg, aes(y = class, fill = drv)) +
  geom_bar()
```

`position = "fill"` will generate a bar plot with bars of equal length and the stacks in each bar will show the proportion of `drv` for that particular `class`.

```{r}
ggplot(mpg, aes(y = class, fill = drv)) +
  geom_bar(position = "fill")
```

If you want to show percentages instead of proportions on the x-axis, you can define this in `scale_x_continuous()` with `scales::label_percent()`.

```{r}
ggplot(mpg, aes(y = class, fill = drv)) +
  geom_bar(position = "fill") +
  scale_x_continuous(name = "percentage", labels = scales::label_percent(accuracy = 1))
```

</details>

### How can I create a stacked bar plot based on data from a contingency table of to categorical variables?

First reshape the data (e.g. with `tidyr::pivot_longer()`) so that there is one row per each combination of the levels of the categorical variables, then use `geom_col()` to draw the bars.

<details>

<summary>See example</summary>

Suppose you have the following data from an opinion poll, where the numbers in the cells represent the number of responses for each party/opinion combination.

```{r}
poll <- tribble(
  ~party,       ~agree, ~disagree, ~no_opinion,
  "Democrat",    20,    30,        20,
  "Republican",  15,    20,        10,
  "Independent", 10,    5,         0
)
```

You can first pivot the data longer to obtain a data frame with one row per party/opinion combination and a new column, `n`, for the number of responses that fall into that category.

```{r}
poll_longer <- poll %>%
  pivot_longer(
    cols = -party,
    names_to = "opinion",
    values_to = "n"
  )

poll_longer
```

Then, you can pass this result to `ggplot()` and create a bar for each `party` on the `y` (or `x`, if you prefer vertical bars) axis and fill the bars in with number of responses for each `opinion`.

```{r}
ggplot(poll_longer, aes(y = party, fill = opinion, x = n)) +
  geom_col()
```

To plot proportions (relative frequencies) instead of counts, use `position = "fill"` in `geom_col()`.

```{r}
ggplot(poll_longer, aes(y = party, fill = opinion, x = n)) +
  geom_col(position = "fill") +
  xlab("proportion")
```

</details>

### How can I make a grouped bar plot?

Map the variable you want to group by to the `x` or `y` `aes`thetic, map the variable you want to color the vars by to the `fill` aesthetic, and set `position = "dodge"` in `geom_bar()`.

<details>

<summary>See example</summary>

Suppose you have data from a survey with three questions, where respondents select "Agree" or "Disagree" for each question.

```{r}
survey <- tibble::tribble(
  ~respondent, ~q1,        ~q2,        ~q3,
  1,           "Agree",    "Agree",    "Disagree",
  2,           "Disagree", "Agree",    "Disagree",
  3,           "Agree",    "Agree",    "Disagree",
  4,           "Disagree", "Disagree", "Agree"
)
```

You'll first want to reshape these data so that each row represents a respondent / question pair.
You can do this with `tidyr::pivot_longer()`.
Then, pass the resulting longer data frame to `ggplot()` group responses for each question together.

```{r}
survey %>%
  tidyr::pivot_longer(
    cols = -respondent,
    names_to = "question",
    values_to = "response"
  ) %>%
  ggplot(aes(x = question, fill = response)) +
  geom_bar(position = "dodge")
```

</details>

### How can I make a bar plot of group means?

Either calculate the group means first and use `geom_col()` to draw the bars or let ggplot2 calculate the means with `stat_summary()` with `fun = "mean"` and `geom = "bar"`.

<details>

<summary>See example</summary>

One option for calculating group means is using `dplyr::group_by()` followed by `dplyr::summarise()`.
Then, you can pass the resulting data frame to `ggplot()` and plot bars using `geom_col()`.

```{r}
mpg %>%
  group_by(drv) %>%
  summarise(mean_hwy = mean(hwy)) %>%
  ggplot(aes(x = drv, y = mean_hwy)) +
  geom_col()
```

Alternatively, you can use `stat_summary()` to let ggplot2 calculate and plot the means.

```{r}
ggplot(mpg, aes(x = drv, y = hwy)) +
  stat_summary(fun = "mean", geom = "bar")
```

</details>

## Axes and axis limits

### Why do the bars on my plot disappear when I specify an axis range with `ylim()`? How can I get the bars to show up within a given axis range?

`ylim()` is a shortcut for supplying the `limits` argument to individual scales.
When either of these is set, any values outside the limits specified are replaced with `NA`.
Since the bars naturally start at `y = 0`, replacing part of the bars with `NA`s results in the bars entirely disappearing from the plot.
For changing axis limits without dropping data observations, set limits in `coord_cartesian()` instead.
Also note that this will result in a deceiving bar plot, which should be avoided in general.

<details>

<summary>See example</summary>

In the following plot the y-axis is limited to 20 to 120, and hence the bars are not showing up.

```{r}
ggplot(mpg, aes(x = drv)) +
  geom_bar() +
  ylim(c(20, 120))
```

In order to obtain a bar plot with limited y-axis, you need to instead set the limits in `coord_cartesian()`.

```{r}
ggplot(mpg, aes(x = drv)) +
  geom_bar() +
  coord_cartesian(ylim = c(20,110))
```

This is, indeed, a deceiving plot.
If you're using a bar plot to display values that could not take the value of 0, you might choose a different geom instead.
For example, if you have the following data and plot.

```{r}
df <- tibble::tribble(
  ~x,  ~y,
  "A", 1050,
  "B", 1100,
  "C", 1150
)

ggplot(df, aes(x = x, y = y)) +
  geom_col()
```

Also suppose that you want to cut off the bars at `y = 1000` since you know that the variable you're plotting cannot take a value less than 1000, you might use `geom_point()` instead.

```{r}
# don't do this
ggplot(df, aes(x = x, y = y)) +
  geom_col() +
  coord_cartesian(ylim = c(1000, 1150))

# do this
ggplot(df, aes(x = x, y = y)) +
  geom_point(size = 3)
```

</details>
